Certifying, incremental type checking

### 2012-06-15 doc/papers/LFMTP2012/paper.tex
paper: finished!

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
a typing derivation out of program $M$; it doesn't have to be
complete, or even sound, and \emph{(ii)} a trusted and fast kernel
$\fcheck{\md}$ to verify these derivations \wrt\ a given set of
inference rules $\Sigma$. This is a \emph{certifying} scheme (see
\cite{leroy2006compcert} for a discussion on approaches to reliable
compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$, but the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or return an
incorrect derivation. Explicitly generating the derivation in the
declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

Several generic methods have been devised for turning a batch program
into an incremental one
\cite{pugh1989incremental,acar2003selective}. Unfortunately, they
suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

@@

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse %TODO le nom?
  to refer to already-computed values;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated incremental certificates
  against a \emph{repository} of already constructed and checked
  derivations.
\end{itemize}

@@

% \begin{figure}
%   \centering
% %  \small
%   \begin{mathpar}
%     \infer{ }{\jts{\cst o} {\even}}

%     \infer{\jts M \odd} {\jts {\s M} \even}

%     \infer{\jts M \even} {\jts {\s M} \odd}

%     \infer{\jts M \nat} {\jts {\s M} \nat}

%     \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}

%     \infer{\infer*{}{\mbox{$
%           \begin{array}{c}
%             [\jts {\var x} A] \\[-0.3em]
%             \vdots\\
%             \jts M B
%           \end{array}$
%         }}} {\jts {\lam x M} {A\to B}}

%     \infer{\jts M A \and \jsub A B}{\jts M B}

%     \infer{\jts M \nat
%       \and
%       \jts N A
%       \and
%       \infer*{}{\mbox{$
%         \begin{array}{c}
%           [\jts {\var x} \nat] \quad
%           [\jts {\var y} A] \\[-0.3em]
%           \vdots \\
%           \jts P A
%         \end{array}
%         $}}
%     }{\jts {\recb M N x y P} A}

%     \infer{ }{\jsub A A}

%     \infer{ }{\jsub \even\nat}

%     \infer{ }{\jsub \odd\nat}

%     \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
%   \end{mathpar}
%   \caption{System \sysname{T_{<:}}, declaratively}
%   \label{fig:tsub}
% \end{figure}

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on those types. He (or
she) wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but \sysname{T_{<:}} is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's
provided incremental verifier. As a first attempt, he wants to verify
$P_1 = \recb{\s\z}{\s\z} x y {\s {\var x}}$: He sends the command
$\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
well-formed program and tries to construct a valid typing derivation
for its argument in the current, empty context. The system succeeds
and returns a typing derivation $\md_1$ of \/ $\vdash P_1 :
\cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates this derivation, it inserts a coercion $\cst{odd}
\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

Let us call $\md_5$ the subderivation generated for the definition of
$\var{add}$, $\md_6$ for the arguments
$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here. $\md_2$ was deriving $\vdash\s\z:\cst{nat}$ (with a
coercion since it was used as an argument to $\var{add}$); the system
thus infer the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.

@@

%TODO figure?

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subterms of the output $\md_i$ back into
subterms of the input $P_i$ is a well-known technique for handling
changes in the input: combined with a way to automatically recognize
already-computed input, it is \emph{memoization}, a general-purpose
technique for incremental computation that has been extensively
studied and combined with other techniques
\cite{pugh1989incremental,acar2003selective}. However, it is best
understood in the case of purely first-order data, and to our best
knowledge remains unexplored for higher-order structures containing
binders, like programs and proofs.

@@

Earlier, the teacher is preparing the tool for the students. He (or
she) provides the abstract syntax, the typing rules and the untrusted
typing algorithm in order to build the incremental type checker. How
should he write the algorithm? We want it to return a derivation, and
take advantage of a higher-order representation of the syntax, so it
shouldn't be necessary to thread any explicit environment. The
definition should then start with:

@@

\noindent
Function $\finfer{}$ takes a term $M$ and produces the pair of an
inferred type $A$ and a derivation that $M$ has type $A$. Its code
will be untyped, but the teacher provides this type so that the
checker can verify that each call is fed and returns values of the
right type: it is its \emph{contract} so to say. He first decomposes
term $M$ and treats the easy cases of $\cst{o}$ and $\cst{s}$:

@@

\noindent
We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types;
patterns don't have to be exhaustive since we are in an untrusted
setting; for conciseness, we use case failure for signaling typing
error. There are three cases for $\s{M}$ in the system, corresponding
to the three rules in the declarative system. For the application
case, we rely on the well-known property that we may need to insert a
coercion from the actual type of the argument to its expected type:

@@

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\ \prd A
{\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error. Notice the slight optimization to the size of the derivation:
if the types $A_1$ and $A_2$ are the same syntactically, we don't
generate a subtyping rule. This is done by matching on the returned
derivation $\md_\leq$.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$, otherwise the check that the
returned open derivation $\md$ is correct would fail ($x$ is
unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
and its $\syntax{let}$ counterpart ($\match M \ldots$ being syntactic
sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
call, the variable case $\var x$ would fail since we would have no way
of telling its type anymore. Thus we substitute all occurrences of
$\var x$ with the \emph{output} of function $\finfer{}$ on it (since
it is known at this point), \ie\ a pair of $A$ and $\md_1$. For this
substitution to be well-typed, we now need to coerce this output to a
valid input (a term). We introduce an operator which is in some sense
the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
type and a derivation to the term it types.

@@

At this point, we get back $B$ and $\md$ from the recursive call. We
know that $B$ can't depend on $\md_x$ (because a System
\sysname{T_{<:}} type is always closed), but $\md$ can: it is not
\emph{per se} a hypothetical derivation but an \emph{open} derivation
where a special judgement named $\md_x$ can occur. To return the
complete derivation of $\vdash\tlam A x M : A \to B$, we must then
close, or ``hypothesize'' $\md$ by $\md_x$, and then apply rule
\rulename{Lam}:

@@

Remains one problem with subtyping: terms $N : A_N$ and $P : A_P$
should be coercible to the same type $A$, but $P$ should be typed
under $\vdash y:A$, which is not known before typing $P$. A trick to
break the loop is to iterate the typing of $P$ with more and more
general types, until we reach a fixpoint (we reach a fixpoint if there
is a unique more general type); in our particular case of subtyping
relation, this means iterating only twice: we will then attain
\cst{nat}.

@@

\noindent
Again, we use a new auxiliary function
$$\sqcap\ :\ \prd{A}{\cst{tp}} \prd{B}{\cst{tp}} \sig C {\cst{tp}}
(\vdash A\leq C )\times(\vdash B\leq C) = \ldots$$ computing the sup
of two types $A$ and $B$, and returning proofs of of subtyping. Note
that we didn't perform here the optimization seen in the application
case.

The type-checker is now done. There is no case for variables, as we
won't ever meet this case: we already substituted all variables with
their result derivation. Evaluation only need to make sure that in
this case, the derivation will be returned, \ie\ that equation
$\finfer{\fget{\pair A \md}} = \pair A \md$ holds.

Note that $\fget{}$ can be derived automatically from the type of
$\finfer{}$: it is its inverse. It has definition:

\begin{mathleft}
  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \lamd M \lamd A \lamd \md M
\end{mathleft}

\noindent
where argument $M$ is implicit.

@@

We exploit here an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
won't return ill-typed values, but arguments and results are checked
at run-time at the boundaries of defined functions. This allows to
omit formal justifications, like the fact that we need to iterate
twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
course this incurs on the performance.

The idea of untrustedly generating certificates verified \emph{a
  posteriori} is very wide-spread in \sysname{LCF}-style proof
assistants: during manual proof search, the interaction is driven by
\emph{tactics}: %TODO cite
they are untrusted programs generating pieces of proofs, assembled by
other, higher-order tactics; verification of the generated proofs is
done when the evaluation of all tactics is complete. The differences
with our mechanism are two-fold: first, tactics usually take one (or
several) \emph{goal}(s) as input, \ie\ holes in a proof, whereas our
functions can take any term; secondly, proof-check is usually done at
the end of the evaluation, which can lead to less tractable bugs; by
making evaluation and checking mutually recursive, we choose to detect
errors earlier in these witnesses: exactly at the point where they are
generated.

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special term construct
$\{x:A\}$ denoting free variables is added to the syntax of terms, and
rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
% TODO: vérif ce paragraphe après avoir parlé de fcts inv
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:comput}).

% TODO cite lee2007mechanizing

@@

% TODO explicit LF
% TODO lambda-lifting

@@

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995lambda}. We differ the discussion of
function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
object containing no function symbols. The definition of hereditary
substitution $\gsubst M \sigma$ is skipped and can be found in
\cite{pfenning2007term} (and extended to metavariables as
\cite{nanevski2008contextual}).

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), as in \eg\ \cite{pierce2000local}: canonical objects
are checked against a given type, whereas we can synthesize the type
of atomic objects; the coercion from the first grammatical category to
the second is where we verify equality between synthesized and checked
types.

>>>>

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to take only the result of the algorithm into
consideration. But let us choose another path for a minute:
derivations in the declarative systems, if they cannot always be
easily inferred, can be \emph{verified} if provided by an oracle. If
the oracle is wrong, we can tell cheaply by verifying the witness
typing derivation it provides. Let us then imagine a type-checker to
be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
a typing derivation out of program $M$; it does not have to be
complete, or even sound, and \emph{(ii)} a trusted and fast kernel
$\fcheck{\md}$ to verify these derivations \wrt\ a given set of
inference rules $\Sigma$. This is a \emph{certifying} scheme (see
\cite{leroy2006compcert} for a discussion on approaches to reliable
compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
\emph{is} a derivation $\md$ for $M$, but the converse is not
necessarily true: function $\finfer{}\!\!$ could fail, or return an
incorrect derivation. Explicitly generating the derivation in the
declarative system has the following advantages over the first
approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item compilers often need to propagate type information further in
  the chain of transformation; the typing derivation produced by
  $\finfer{}$ being explicit, it could constitute the data structure
  being exploited.
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\end{itemize}
\ldots\ but there is more:

@@

Several generic methods have been devised for turning a batch program
into an incremental one \cite{acar2006,carlsson2002}. Unfortunately,
they suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

@@

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse
  to refer to already-computed values;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated incremental certificates
  against a \emph{repository} of already constructed and checked
  derivations.
\end{itemize}

@@

A student is learning about System \sysname {T_{<:}} --- a simply
typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
elimination ($\recb M N x y P$), and subtyping on these types (for
formal details, see appendix). He (or she) wants to write a program
computing $(2+2)\times 3$ in that system. He knows that sometimes the
arguments of $\cst{rec}$ are quite hard to get correct, but
\sysname{T_{<:}} is a typed language; he wants to take advantage of
this to build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1 =
\recb{\s\z}{\s\z} x y {\s {\var x}}$: he sends the command
$\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
well-formed program and tries to construct a valid typing derivation
for its argument in the current, empty context. The system succeeds
and returns a typing derivation $\md_1$ of \/ $\vdash P_1 :
\cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$. He
could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program: this is a
\emph{delta}. Derivation $\md_2$ used to live in the empty context, we
are then free to use it in the new context where
$\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T
enjoys the \emph{weakening} property; $\md_4$ depending on a
derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole with
the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates this derivation, it inserts a coercion $\cst{odd}
\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them.

Let us call $\md_5$ the subderivation generated for the definition of
$\var{add}$, $\md_6$ for the arguments
$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the application
$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ does not depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\fget{\md_5}}
  \letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\fget{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so we just have to
use it here. $\md_2$ was deriving $\vdash\s\z:\cst{nat}$ (with a
coercion since it was used as an argument to $\var{add}$); the system
thus infers the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.

@@

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subterms of the output $\md_i$ back into
subterms of the input $P_i$ is a well-known technique for handling
changes in the input: combined with a way to automatically recognize
already-computed input, it is \emph{memoization}, a general-purpose
technique for incremental computation that has been extensively
studied and combined with other techniques \cite{acar2006}. However,
it is best understood in the case of purely first-order data, and to
our best knowledge remains unexplored for higher-order structures
containing binders, like programs and proofs.

@@

Earlier, the teacher is preparing the tool for the students. He (or
she) provides the abstract syntax, the typing rules and the untrusted
typing algorithm in order to build the incremental type checker. How
should he write the algorithm? We want it to return a derivation, and
take advantage of a higher-order representation of the syntax, so it
should not be necessary to thread any explicit environment. The
definition should then start with:

@@

\noindent
Function $\finfer{}$ takes a term $M$ and produces the pair of an
inferred type $A$ and a derivation that $M$ has type $A$. Its code
will be untyped, but the teacher provides this type so that the
checker can verify that each call is fed and returns values of the
right type: it is its \emph{contract} so to say. He first decomposes
term $M$ and treats the easy cases of $\cst{o}$ and $\cst{s}$. The
code continues with:

@@

\noindent
We write $\pair{\cdot}{\cdot}$ for the constructor of $\Sigma$-types.
Patterns do not have to be exhaustive since we are in an untrusted
setting. Fr conciseness, we use case failure for signaling typing
error. There are three cases for $\s{M}$ in the system, corresponding
to the three rules in the declarative system. For the application
case, we rely on the well-known property that we may need to insert a
coercion from the actual type of the argument to its expected type:

@@

\noindent
The $\syntax{let}$ construct is just syntactic sugar for a one-branch
$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\ \prd A
{\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
construct a subtyping derivation for its argument, or failing with an
error. Notice the slight optimization to the size of the derivation:
if the types $A_1$ and $A_2$ are the same syntactically, we do not
generate a subtyping rule. This is done by matching on the returned
derivation $\md_\leq$ in the case of the \textsc{Refl} rule.

Then comes the abstraction case $\tlam x A M$. The difficulty is that
the recursive call $\finfer{M}$ must be done in an enlarged
environment where $\vdash x : A$. Otherwise the check that the
returned open derivation $\md$ is correct would fail ($x$ is
unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
($\match M \ldots$ being syntactic sugar for $\matchin M \envnil
\ldots$). Even then, in the recursive call, we would have no way of
synthesizing a variable $\var x$'s type anymore. Thus we substitute
all occurrences of $\var x$ with the \emph{output} of function
$\finfer{}$ on it (since it is known at this point), \ie\ a pair of
$A$ and $\md_1$. For this substitution to be well-typed, we now need
to coerce this output to a valid input (a term). We introduce an
operator which is in some sense the \emph{inverse} of $\finfer{}$:
operator $\fget{}$ maps a pair of a type and a derivation to the term
it types.

@@

At this point, we get back $B$ and $\md$ from the recursive call. We
know that $B$ cannot depend on $\md_x$ (because a System
\sysname{T_{<:}} type is always closed), but $\md$ can: it is not
\emph{per se} a hypothetical derivation but an \emph{open} derivation
where a special judgement named $\md_x$ can occur. To return the
complete derivation of $\vdash\tlam x A M : A \to B$, we must then
close, or ``hypothesize'' $\md$ by $\md_x$, and then apply rule
\rulename{Lam}:

@@

One problem remains with subtyping: terms $N : A_N$ and $P : A_P$
should be coercible to the same type $A$, but $P$ should be typed
under $\vdash y:A$, which is not known before typing $P$. A trick to
break the loop is to iterate the typing of $P$ with a more and more
general type, until we reach a fixpoint. We reach a fixpoint if there
is a unique more general type, in our particular case of subtyping
relation, this means iterating only twice: we will then attain
\cst{nat}.

@@

\noindent
Again, we use a new auxiliary function
$$\sqcap\ :\ \prd{A}{\cst{tp}} \prd{B}{\cst{tp}} \sig C {\cst{tp}}
(\vdash A\leq C )\times(\vdash B\leq C) = \ldots$$ computing the sup
of two types $A$ and $B$, and returning proofs of subtyping. Note that
we could have performed the optimization seen in the application case
as well.

The type-checker is now done. There is no case for variables, as we
will not ever meet this case: we already substituted all variables
with their result derivation. Evaluation only needs to make sure that
in this case, the derivation will be returned, \ie\ that equation
$\finfer{\fget{\pair A \md}} = \pair A \md$ holds.

Note that $\fget{}$ can be derived automatically from the type of
$\finfer{}$: we will call it its inverse, since it maps the return
type to the arguments. It has definition:

\begin{mathleft}
  \fget{}\ :\ \prd{M}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
    M:A)} \cst{tm} =
  \lamd M \lamd A \lamd \md M
\end{mathleft}

\noindent
where argument $M$ is left implicit (it can be inferred).

@@

We exploit an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
will not return ill-typed values, but arguments and results are
checked at run-time at the boundaries of defined functions. This
allows to omit formal justifications, like the fact that we need to
iterate twice the typing of $P$ in $\recb M N x y P$, or that $B$
cannot depend on $\md_x$ in the $\lambda$ case, while detecting errors
early.

The idea of untrustedly generating certificates verified \emph{a
  posteriori} is very wide-spread in \sysname{LCF}-style proof
assistants: during manual proof search, the interaction is driven by
\emph{tactics}: they are untrusted programs generating pieces of
proofs, assembled by other, higher-order tactics; verification of the
generated proofs is done when the evaluation of all tactics is
complete. The differences with our mechanism are two-fold: first,
tactics usually take one (or several) \emph{goal}(s) as input, \ie\
holes in a proof, whereas our functions can take any term; secondly,
proof-check is usually done at the end of the evaluation, which can
lead to less tractable bugs; by making evaluation and checking
mutually recursive, we choose to detect errors earlier in these
witnesses: exactly at the point where they are generated.

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special term construct
$\{x:A\}$ denoting free variables is added to the syntax of terms, and
rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:comput}).

@@

@@

In this variant, only $\eta$-long and $\beta$-normal objects have an
existence. Canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
(\ie\ to a \emph{spine} of objects) as opposed to the more standard
binary application: this eases the definition of substitution and
$\eta$-long forms, and corresponds to a focused sequent calculus known
as \sysname{LJT}~\cite{herbelin1995lambda}. We differ the discussion
of function symbols $\fct f$ to section \ref{sec:comput}. A
\emph{value} is an object containing no function symbols. The
definition of hereditary substitution $\gsubst M \sigma$ is skipped
and can be found in the appendix.

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), as in \eg\ \cite{pierce2000local}. Canonical objects
are checked against a given type, whereas we can synthesize the type
of atomic objects; the coercion from canonical to atomic objects
triggers the equality check between synthesized and checked types.

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A_1\to A_2$ for $\prd x {A_1} {A_2}$ when $\var
x\notin \FV{A_2}$, $\meta X$ for $\smeta X \msubstnil$, and adopt
indifferently list, map or set notations for $\sigma$ and
$\Gamma$. The identity substitution $\msubstid{\Gamma}$ is a notation
for $\{\msubst x {\var x}\ |\ \var x \in\dom\Gamma\}$. Metavariables
$\smeta X \sigma$ stand for open objects with free variables in
$\dom{\sigma}$. To each use of them is attached a substitution
$\sigma$ defining these free variables. We write $\FMV{M}$ for the set
of metavariables in $M$ and call an object $M$ \emph{metaclosed} if
$\FMV{M} = \emptyset$.
